---
title: "Biostat 200C Homework 5"
subtitle: Due 11:59PM June 2nd
author: Yenlin Lai
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, cache = FALSE)
library(tidyverse)
library(faraway)
```

## Q1. Doctor visits in Australia, ELMR Exercise 13.4 

```{r}
data(dvisits)
help("dvisits")
```

The `dvisits` data comes from the Australian Health Survey of 1977-78 and consist of 5190 single adults where young and old have been oversampled. Use `help("dvisits")` to check the variables.

(a) Build a generalized additive model with `doctorco` as the response and `sex`, `age`,`agesq`, `income`, `levyplus`, `freepoor`, `freerepa`, `illness`, `actdays`, `hscore`, `chcond1` and `chcond2` as possible predictor variables. Select an appropriate size for your model. (Hint. fit a simpler model first and check some marginal plots.)

**Solution:**
Check the marginal plots first. From the marginal plots below, we can see that some of the variables (`sex`, `levyplus`, `freepoor`, `freerepa`, `chcond1` and `chcond2`) cannot fit a spline/smooth in the generalized additive model.
```{r}
predlist <- setdiff(colnames(dvisits), 
                   c("doctorco", "nondocco", "hospadmi", "hospdays",
                     "medicine", "prescrib", "nonpresc"))
dvisits %>% 
  pivot_longer(all_of(predlist), 
               names_to = "predictor", values_to = "value") %>%
  ggplot() +
  geom_point(aes(x = value, y = doctorco)) +
  facet_wrap(~ predictor, scales = "free") +
  geom_smooth(aes(x = value, y = doctorco)) 
```

We then fit a simpler model - generalized linear model (poisson regression model).

```{r}
lmod = glm(doctorco ~ sex + age + agesq + income + levyplus + freepoor + 
            freerepa + illness + actdays + hscore + chcond1 + chcond2
          ,data = dvisits, family = "poisson")
summary(lmod)
halfnorm(cooks.distance(lmod))
```

The influential plot suggests that case 115 is an outlier. We remove this observation for the following analyses.

```{r}
dvisits2 <- dvisits[-115,]
lmod2 = glm(doctorco ~ sex + age + agesq + income + levyplus + freepoor + 
            freerepa + illness + actdays + hscore + chcond1 + chcond2
          ,data = dvisits2, family = "poisson")
summary(lmod2)
halfnorm(cooks.distance(lmod2))
```

It seems that there is no outliers in the sub-setting data. Use the package `gam` to fit the generalized additive model, with poisson distribution.

```{r}
library(gam)
amod <- gam(doctorco ~ sex + s(age) + s(agesq) + s(income) + levyplus + freepoor
            + freerepa + s(illness, 6) + s(actdays) + s(hscore) + chcond1
            + chcond2, data = dvisits2, scale = -1, family = "poisson")
summary(amod)
```

We use stepwise method to choose the optimal model.

```{r}
scope_list = list(
  "age" = ~1 + s(age) + s(age, 2) + s(age, 3),
  "agesq" = ~1 + s(agesq) + s(agesq, 2) + s(agesq, 3),
  "income" = ~1 + s(income) + s(income, 2) + s(income, 3),
  "illness" = ~1 + s(illness) + s(illness, 2) + s(illness, 3) + s(illness, 4),
  "actdays" = ~1 + s(actdays) + s(actdays, 2) + s(actdays, 3),
  "hscore" = ~1 + s(hscore) + s(hscore, 2) + s(hscore, 3),
  "sex" = ~1 + sex,
  "levyplus" = ~1 + levyplus,
  "freepoor" = ~1 + freepoor,
  "freerepa" = ~1 + freerepa,
  "chcond1" = ~1 + chcond1,
  "chcond2" = ~1 + chcond2)
step.Gam(amod, scope_list)

amod2 <- gam(doctorco ~ s(illness, 6) + s(age, 2) + s(income, 2) + s(actdays) +
               s(hscore, 2) + sex + freepoor, family = "poisson", 
             data = dvisits2, scale = -1)
summary(amod2)
```

The final model is `doctorco ~ s(illness, 6) + s(age, 2) + s(income, 2) + s(actdays) + s(hscore, 2) + sex + freepoor`, and all the variables are significant. We then fit the final model again to the package `mgcv` to plot the transformations on the predictors identified by the additive model. 

```{r}
amod3 <- mgcv::gam(doctorco ~ s(illness, k = 6) + s(age, k = 2) + 
                     s(income, k = 2) + s(actdays) + s(hscore, k = 2) + sex +
                     freepoor, family = "poisson", data = dvisits2, scale = -1)
summary(amod3)
plot(amod3, residuals = TRUE, page = 1)
```

We can see that the confidence bands do not highly overlap with `s() = 0` for those variables, thus, these variables are significant.

(b) Check the diagnostics.

**Solution:**
```{r}
library(mgcv)
halfnorm(cooks.distance(amod3))
gam.check(amod3)
```

From the output, the small p-value indicates the residuals are not randomly distributed. Since we have all the p-values greater than 0.05, we can say the residuals for each variables are randomly distributed.

From the diagnostics plots above, the Q-Q plot do not fit well, which means the data may not meet the normality. For the `histogram of residuals`, the distribution should be bell-shaped and centered at 0, and our model seems to not be the best. We also expect to see the dots are evenly distributed and centered at 0 in the `resid vs. linear pred.` plot, and the dots should be lined up in the `response vs. fitted values` plot; therefore, the model does not fit really well.


(c) What sort of person would be predicted to visit the doctor the most under your selected model?

**Solution:**
```{r}
summary(amod3)
```

The estimated coefficients for `sex` is 0.156, which means female would be predicted to visit the doctor more than male.
The estimated coefficients for `freepoor` is -0.73, which means people who covered by government because low income, recent immigrant, unemployed would be predicted to visit the doctor more.

```{r}
plot(amod3, page = 1)
```
From the plots shown above, people whose `illness` are 2 to 5, `age` are 72, `income` are less than 200, `actdays` are 11.5, and `hscore` are 12 have a highest probability to visit the doctor the most.

Combine the two results, and we get people who is female and covered by government because low income, recent immigrant, unemployed, and whose `illness` are 2 to 5, `age` are 72, `income` are less than 200, `actdays` are 11.5, and `hscore` are 12 would be predicted to visit the doctor the most.


(d) For the last person in the dataset, compute the predicted probability distribution for their visits to the doctor, i.e., give the probability they visit 0,1, 2, etc. times.

**Solution:**
```{r}
predict(amod3, dvisits2[5189,], type = "response")
```

The parameter of poisson distribution for the last person in the dataset is 0.1037571.

```{r}
dpois(0, 0.1037571)
dpois(1, 0.1037571)
dpois(2, 0.1037571)
dpois(3, 0.1037571)
dpois(4, 0.1037571)
dpois(5, 0.1037571)
dpois(6, 0.1037571)
dpois(7, 0.1037571)
dpois(8, 0.1037571)
dpois(9, 0.1037571)
```

The probability they visit 0 time is 0.9014442;
the probability they visit 1 time is 0.09353124;
the probability they visit 2 times is 0.004852265;
the probability they visit 3 times is 0.000167819;
the probabilities they visit 4,5,6,7,8, and 9 times are almost 0.
